Equalizer Programming Guide

### 2008-01-16 Developer/ProgrammingGuide/paper.tex
Update Programmers Guide
[ ] May break build
[ ] Breaks existing applications
[ ] Bugfix
[ ] New Feature
[ ] Cleanup
[ ] Optimization
[x] Documentation

\date{
  \vspace{2cm}
  \begin{figure}[ht]
    \hspace{-4.5cm}\includegraphics[width=22cm]{images/teaser.pdf}
  \end{figure}
  \vspace{1cm}
%  \textbf{RELEASE CANDIDATE}\\[\medskipamount]
  Version 1.1.8 for Equalizer 0.4+svn\\[\medskipamount]
  \today
}

@@

\clearpage
\thispagestyle{empty}
\tableofcontents
\vfill{\center\begin{tabularx}{\textwidth}{|l|l|X|}
    \hline
    \bf Rev & \bf Date     & \bf Changes \\
    \hline
    1.0     & Oct 28, 2007 & Initial Version for Equalizer 0.4\\
    1.0.1   & Nov 06, 2007 & Fixed Typo\\
    1.1     & Nov 12, 2007 & Adapted distributed objects\\
    1.1.1   & Nov 16, 2007 & Minor improvements based on lulu.com print\\
    1.1.2   & Nov 19, 2007 & Minor improvements based on feedback\\
    1.1.3   & Nov 23, 2007 & Add OpenGL extension handling\\
    1.1.4   & Nov 26, 2007 & Changes caused by removal of glX event thread\\
    1.1.5   & Dec 10, 2007 & NodeFactory API changes\\
    1.1.6   & Dec 17, 2007 & Add Scalable Rendering chapter\\
    1.1.7   & Dec 20, 2007 & Per-window object manager changes\\
    1.1.8   & Jan 15, 2007 & Detailed window initialization based on
                             latest refactoring\\
    \hline
  \end{tabularx}}
\thispagestyle{empty}
\clearpage

@@

The main thread is responsible for maintaining the application logic. It
reacts on user events, updates the data model and requests new frames to
be rendered. It drives the whole application, as shown in \fig{fModel}.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/model.pdf}
  {\caption{\small\label{fModel}Simplified execution model}}
\end{wrapfigure}
All rendering threads in the configuration run asynchronously to the
application's main thread. Depending on the configuration's latency,
they can fall $n$ frames behind the last frame finished by the
application thread. A latency of one frame is usually not perceived by
the user, but can increase rendering performance substantially.

@@

\subsubsection{Frustum}
\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/frustra.pdf}
  {\caption{\small\label{fFrustra}Wall and projection parameters}}
\end{wrapfigure}
Compounds have a frustum description to define the physical layout of
the display environment.

The frustum description is inherited by the children, therefore the
frustum is typically defined on the top-most compound.

@@

By default, \textsf{Config::finish\-Fra\-me} also synchronizes the
completion of all local rendering tasks for the current frame. This
facilitates porting of exising rendering codes, since the database does
not have to be multi-buffered. Applications such as \textsf{eqPly}, which
do not need this per-node frame synchronization can disable it, as
explained in \sref{sNodeFrame}.

\begin{wrapfigure}{r}{.618\textwidth}
  \subfigure[]{\label{fSync}
    \includegraphics[width=.25\textwidth]{images/sync}}\hfil
  \subfigure[]{\label{fAsync}
    \includegraphics[width=.25\textwidth]{images/async}}%
  {\caption{\small\label{fSyncAsync}Synchronous and asynchronous execution}}
\end{wrapfigure}
\fig{fSyncAsync} shows the execution of (hypothetical) rendering tasks
without latency (\fig{fSync}) and with a latency of
one frame (\fig{fAsync}). With \textsf{eqPly}, a speedup of 15\% has been
observed on a five-node rendering cluster when using a latency of one frame
instead of no
latency\footnote{\link{http://www.equalizergraphics.com/scalability.html}}. A
latency of one or two frames is normally not perceived by the user.

@@

@@

The config initialization happens in parallel, that is, all config
initialization tasks are transmitted by the server at once and their
completion is synchronized afterwards.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/configInit.pdf}
  {\caption{\small\label{fConfigInit}Config Initialization Sequence}}
\end{wrapfigure}
If the configuration was initialized correctly, the configuration tries
to set up a tracking device for head tracking. Equalizer does not
provide extensive support for tracking devices, as this is an orthogonal
problem to parallel rendering. Tracking device support has already been
solved by a number of implementations\footnote{VRCO Trackd, VRPN, etc.},
which can easily be integrated with Equalizer. The example code in
\textsf{eqPly} provides a reference implementation for the integration
of such a tracking library. \sref{sTracking} provides more background on
head tracking.

@@

The initialization sequence uses multiple, overrideable task
methods. The main task method \textsf{configInit} executes a `child'
task method to create the drawable and context. The child task method
depends on the window system of the pipe. The default implementations of
\textsf{configInitGLX}, \textsf{configInitWGL} or \textsf{configInitAGL}
create an on-screen window using OS-specific methods. If the OpenGL
drawable and context were created successfully, \textsf{configInit} calls
\textsf{configInitGL}, which performs the generic OpenGL
state initialization. The default implementation sets up some typical
OpenGL state, e.g., it enables the depth test.

@@

\begin{wrapfigure}{r}{.382\textwidth}
  \includegraphics[width=.382\textwidth]{images/windowInit.pdf}
  {\caption{\small\label{fWindowInit}Window Initialization}}
\vspace{-4ex}\end{wrapfigure}
The window-system specific initialization takes into account various
attributes set in the configuration file. Attributes include the size of
various frame buffer attachments (color, alpha, depth, stencil) as well
as other framebuffer properties, such as quad-buffered stereo,
doublebuffering, fullscreen mode and window decorations. Some of the
attributes, such as stereo, doublebuffer and stencil can be set to
\textsf{eq::AUTO}, in which case Equalizer will test for their availability and 
enable them if possible.

The \textsf{eqPly} window initialization function first calls
\textsf{eq::Window::configInit} to use the generic window setup. If that
was successful, it initializes a state object and an overlay logo:
\vspace{4ex}

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \vspace{-2.5ex}\includegraphics[width=.618\textwidth]{images/render.pdf}
  {\caption{\small\label{fRender}Main Rendering Loop}\vspace{-2ex}}
\end{wrapfigure}
The model data is spatially organized in an 3-dimensional
kD-tree\footnote{See also \link{http://en.wikipedia.org/wiki/Kd-tree}}
for efficient view frustum culling. When the model is loaded by
\textsf{Node::config\-Init}, it is preprocessed into the kD-tree and each
node of the tree gets a database range assigned. The root node
has the range [0, 1], its left child [0, 0.5] and its right child [0.5,
1], and so on for all nodes in the tree. The preprocessed model is saved
in a binary format for accelerating subsequent use.

@@

In order to provide a motivation for the design of the image compositing
API, the direct send parallel compositing algorithm is introduced in this
section.

>>>>

\date{
  \vspace{2cm}
  \begin{figure}[ht]
    \hspace{-4.5cm}\includegraphics[width=22cm]{images/teaser.pdf}
  \end{figure}
  \vspace{1cm}
%  \textbf{RELEASE CANDIDATE}\\[\medskipamount]
  Version 1.1.9 for Equalizer 0.4+svn\\[\medskipamount]
  \today
}

@@

\clearpage
\thispagestyle{empty}
\tableofcontents
\vfill{\center\begin{tabularx}{\textwidth}{|l|l|X|}
    \hline
    \bf Rev & \bf Date     & \bf Changes \\
    \hline
    1.0     & Oct 28, 2007 & Initial Version for Equalizer 0.4\\
    1.0.1   & Nov 06, 2007 & Fixed Typo\\
    1.1     & Nov 12, 2007 & Adapted distributed objects\\
    1.1.1   & Nov 16, 2007 & Minor improvements based on lulu.com print\\
    1.1.2   & Nov 19, 2007 & Minor improvements based on feedback\\
    1.1.3   & Nov 23, 2007 & Add OpenGL extension handling\\
    1.1.4   & Nov 26, 2007 & Changes caused by removal of glX event thread\\
    1.1.5   & Dec 10, 2007 & NodeFactory API changes\\
    1.1.6   & Dec 17, 2007 & Add Scalable Rendering chapter\\
    1.1.7   & Dec 20, 2007 & Per-window object manager changes\\
    1.1.8   & Jan 15, 2007 & Detailed window initialization based on
                             latest refactoring\\
    1.1.9   & Jan 16, 2007 & Detail distributed object usage\\
    \hline
  \end{tabularx}}
\thispagestyle{empty}
\clearpage

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/model.pdf}
  {\caption{\small\label{fModel}Simplified execution model}}
\end{wrapfigure}
The main thread is responsible for maintaining the application logic. It
reacts on user events, updates the data model and requests new frames to
be rendered. It drives the whole application, as shown in \fig{fModel}.

@@

All rendering threads in the configuration run asynchronously to the
application's main thread. Depending on the configuration's latency,
they can fall $n$ frames behind the last frame finished by the
application thread. A latency of one frame is usually not perceived by
the user, but can increase rendering performance substantially.

@@

\subsubsection{Frustum}
Compounds have a frustum description to define the physical layout of
the display environment.

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/frustra.pdf}
  {\caption{\small\label{fFrustra}Wall and projection parameters}}
\end{wrapfigure}
The frustum description is inherited by the children, therefore the
frustum is typically defined on the top-most compound.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \subfigure[]{\label{fSync}
    \includegraphics[width=.25\textwidth]{images/sync}}\hfil
  \subfigure[]{\label{fAsync}
    \includegraphics[width=.25\textwidth]{images/async}}%
  {\caption{\small\label{fSyncAsync}Synchronous and asynchronous execution}}
\end{wrapfigure}
By default, \textsf{Config::finish\-Fra\-me} also synchronizes the
completion of all local rendering tasks for the current frame. This
facilitates porting of exising rendering codes, since the database does
not have to be multi-buffered. Applications such as \textsf{eqPly}, which
do not need this per-node frame synchronization can disable it, as
explained in \sref{sNodeFrame}.

\fig{fSyncAsync} shows the execution of (hypothetical) rendering tasks
without latency (\fig{fSync}) and with a latency of
one frame (\fig{fAsync}). With \textsf{eqPly}, a speedup of 15\% has been
observed on a five-node rendering cluster when using a latency of one frame
instead of no
latency\footnote{\link{http://www.equalizergraphics.com/scalability.html}}. A
latency of one or two frames is normally not perceived by the user.

@@

\subsubsection{Best Practices}

Distributed objects are addressed using session-unique identifiers,
because pointers to other objects can not be distributed directly, since
they have no meaning on remote nodes. The session with which distributed
objects are registered is normally the \textsf{eq::Config}.

The shared data entry point for the render clients is the identifier
passed by the application to \textsf{Config::init}. This identifier
typically contains the identifier of a static distributed object, and is
passed by Equalizer to all \textsf{configInit} task methods. Normally
this initial object is mapped by the \textsf{eq::Node}, and contains
identifiers to other shared data objects.

The distributed data objects referenced by the initial data object are
often versioned objects, to keep them in sync with the rendered
frames. Similar to the initial identifier passed to
\textsf{Config::init}, an object identifier or version can be passed to
\textsf{Config::startFrame}. Equalizer will pass this identifier to all
\textsf{frameStart} task methods. In \textsf{eqPly}, the frame-specific
data, e.g., the global camera data, is versioned. The frame data
identifier is passed in the initial data, and the frame data version is
passed with each start frame request.

\begin{wrapfigure}{r}{.382\textwidth}
  \includegraphics[width=.382\textwidth]{images/objects.pdf}
  {\caption{\small\label{fObjects}Distributed Data Handling}}
\end{wrapfigure}
Applications distributing their scene graph, rather then relying on a
shared filesystem like \textsf{eqPly}, use the frame data also as an
entry point to their scene graph data structure. \fig{fObjects} shows
one possible implementation, where the identifier and version of the
scene graph root are transported using the frame data. The scene graph
root then serializes and de-serializes his immediate children by
transferring their identifier and current version.

The objects are still created by the application, and then registered or
mapped with the session in order to distribute them. When mapping
objects in a hierarchical data structure, their type often has to be
known in order to create them. Equalizer does not currently provide
object typing, this has to be done by the application, either implicitly
in the current implementation context, or by transferring a type
identifier. In \textsf{eqPly}, object typing is implicit since it is
well-defined which object is mapped in which task method.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/configInit.pdf}
  {\caption{\small\label{fConfigInit}Config Initialization Sequence}}
\end{wrapfigure}
The config initialization happens in parallel, that is, all config
initialization tasks are transmitted by the server at once and their
completion is synchronized afterwards.

@@

If the configuration was initialized correctly, the configuration tries
to set up a tracking device for head tracking. Equalizer does not
provide extensive support for tracking devices, as this is an orthogonal
problem to parallel rendering. Tracking device support has already been
solved by a number of implementations\footnote{VRCO Trackd, VRPN, etc.},
which can easily be integrated with Equalizer. The example code in
\textsf{eqPly} provides a reference implementation for the integration
of such a tracking library. \sref{sTracking} provides more background on
head tracking.

@@

\begin{wrapfigure}{r}{.382\textwidth}
  \includegraphics[width=.382\textwidth]{images/windowInit.pdf}
  {\caption{\small\label{fWindowInit}Window Initialization}}
\end{wrapfigure}
The initialization sequence uses multiple, overrideable task
methods. The main task method \textsf{configInit} executes a `child'
task method to create the drawable and context. The child task method
depends on the window system of the pipe. The default implementations of
\textsf{configInitGLX}, \textsf{configInitWGL} or \textsf{configInitAGL}
create an on-screen window using OS-specific methods. If the OpenGL
drawable and context were created successfully, \textsf{configInit} calls
\textsf{configInitGL}, which performs the generic OpenGL
state initialization. The default implementation sets up some typical
OpenGL state, e.g., it enables the depth test.

@@

The window-system specific initialization takes into account various
attributes set in the configuration file. Attributes include the size of
various frame buffer attachments (color, alpha, depth, stencil) as well
as other framebuffer properties, such as quad-buffered stereo,
doublebuffering, fullscreen mode and window decorations. Some of the
attributes, such as stereo, doublebuffer and stencil can be set to
\textsf{eq::AUTO}, in which case Equalizer will test for their availability and 
enable them if possible.

The \textsf{eqPly} window initialization function first calls
\textsf{eq::Window::configInit} to use the generic window setup. If that
was successful, it initializes a state object and an overlay logo:

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/render.pdf}
  {\caption{\small\label{fRender}Main Rendering Loop}}
\end{wrapfigure}
The model data is spatially organized in an 3-dimensional
kD-tree\footnote{See also \link{http://en.wikipedia.org/wiki/Kd-tree}}
for efficient view frustum culling. When the model is loaded by
\textsf{Node::config\-Init}, it is preprocessed into the kD-tree and each
node of the tree gets a database range assigned. The root node
has the range [0, 1], its left child [0, 0.5] and its right child [0.5,
1], and so on for all nodes in the tree. The preprocessed model is saved
in a binary format for accelerating subsequent use.

@@

In order to provide a motivation for the design of the image compositing
API, the direct send parallel compositing algorithm is introduced in this
section. Other parallel compositing algorithms, e.g. binary-swap, can
also be expressed through an Equalizer configuration file.
